---
title: "QMM1002 Case Study  2 [20%]"
author:
- Name Here
- Student Number Here
date: 'Due: April 21, 2023 at 11:59PM'
output:
  html_document: default
  word_document: default
  pdf_document: default
extra_dependencies: xcolor
---


## **<span style="color: #8F2050;">Introduction</span>**

In QMM1001 and QMM1002 all students collected personalized data to analyze. Using RMarkdown to analyze this data, you can show your code and output from R directly in your report. 

In the introduction section, you must describe the variables collected. All students had to collect data on hours studying each day of their first 2 terms of study. We can also identify the program code for each student. 

You can include a list of your variables in an RMarkdown table. The relevant variables for Case Study 2 are: 

Variable | Type
------------- | -------------
Date | Identifier
Hours Studying  | Quantitative
Term (F22, W23) | Categorical
Program (BAPG, CAGC, HAGC) | Categorical

You must also include calculations of summary statistics. This is how you run code in RMarkdown. Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the summary statistics, only the output is shown. You can also set `include = FALSE` to hide all code and output.

**Summary Statistics**

These summary statistics can be used to compare the amount of time you spend studying to other students and to other students by program. 

```{r include=FALSE}
personal<-read.csv(file="PersonalizedData.csv", header=TRUE)
personal.studymean<-mean(personal$Study, na.rm=TRUE)
personal.studysd<-sd(personal$Study, na.rm=TRUE)

combined<-read.csv(file="Combined.csv", header=TRUE, fileEncoding="UTF-8-BOM")
```

Data Set | Mean | Standard Deviation
------------- | ------------- | ------------- 
personal | `r mean(personal$Study, na.rm=TRUE)` | `r sd(personal$Study, na.rm=TRUE)`
All Students | `r mean(combined$Study, na.rm=TRUE)` | `r sd(combined$Study, na.rm=TRUE)`
BAPG | Value here | Value here
CAGC | Value here | Value here
HAGC | Value here | Value here


You will use your own personalized data set and a combined class data set to answer the questions: 

1. Are there differences in the average study times for students in the different analytics streams?
2. Is the distribution of days studied more than 3.13 hours (the average daily study time for students at McGill) the same for students in the different analytics streams (or in other words, independent of program stream)?
3. How does your personal study time change over time?

You must also summarize the methods that will be used throughout Case Study 2 to answer these questions.


## **<span style="color: #8F2050;">Data Analysis</span>**
Don't forget you must include more than 5 examples of your daily activities that help to explain the results of your analysis for full marks in this section!

### Part 1: ANOVA

Place your analysis here.

You can easily write hypotheses like this: 

$H_0: \mu_1 = \mu_2 = ... = \mu_k$

$H_A:$ At least one mean is different

```{r include=FALSE}
#include code within chunks that have hidden output

#Follow instruction on Module 6 Applied Activity Case Study 2 Check Point
#for creating random subsets for the streams 

#use last digits of student number ex. 138572 to get SAME random 
#numbers each time you run your code so your output doesn't change
set.seed(138572)
```

### Part 2: Chi-Square Tests

Place your analysis here.

```{r include=FALSE}
#include code within chunks that have hidden output

#Follow instruction on Module 9 Applied Activity Case Study 2 Check Point
#for creating categorical variable to analyze counts
```

### Part 3: Time Series Analysis

Place your analysis here. 

```{r include=FALSE}
#you must install these packages in the console first using 
#install.packages("zoo")
#install.packages("forecast")
library(zoo)
library(forecast)

personal$Date <- as.Date(personal$Date,format="%m/%d/%Y") #Check your format! Dependening on how you entered dates in Excel, you might have to change it
personal.zoo <- zoo(personal[,c(-1, -2, -4, -5, -6, -7, -8, -9, -10)],personal[,1]) #remove all columns except study time and set dates to index

personal.all <- merge(personal.zoo,zoo(,seq(start(personal.zoo),end(personal.zoo),by="day")), all=TRUE) #include all missing dates

#Make a time series with the longest stretch of dates
personal.ts<-ts(na.contiguous(personal.all))
```


```{r}
#you can then do any of the time series steps with your new data set!
plot.ts(personal.ts, xlab="Days since August 31, 2021", ylab="Hours Studied", main="Time Series Plot")
```

## **<span style="color: #8F2050;">Conclusion</span>**
Summarize your findings for all 3 questions: 

1. Are there differences in the average study times for students in the different analytics streams?
2. Is the distribution of days studied more than 3.13 hours (the average daily study time for students at McGill) the same for students in the different analytics streams (or in other words, independent of program stream)?
3. How does your personal study time change over time?

and explain what you learned about yourself and your daily habits by completing the report. 
